---
title: "Healthcare Big Data Analytics"
subtitle: "Data Cleaning, Recoding, Reshaping, and Missingness"
author: "Ben Gerber, MD, MPH"
date: last-modified
format:
  revealjs:
    theme: dracula
    slide-number: true
    smaller: true
    footer: "Data Cleaning & Wrangling"
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Today's Goals

By the end of this lesson, you will be able to:

::: {.incremental}
- Recognize common **messy patterns** (지저분한) in real-world EHR data
- **Recode** values into clean categories
- **Reshape** data between wide and long formats using `pivot_longer()` / `pivot_wider()`
- Detect **outliers** (이상치) using visual and statistical methods
- Diagnose **patterns of missingness** using R packages
:::

---

## Why This Matters

::: {.incremental}
- Real EHR data is entered by many people and systems over many years — inconsistency is common
- A model or cohort built on unrecoded, unreshaped, or silently-missing data produces **wrong answers with high confidence**
- This is typically **75% of the actual time** spent with real health data
- Cleaning usually happens *before* or *between* joins
:::

---

# Part 1: A Messy Synthetic Dataset

---

## Let's Make a Mess

We'll make a small OMOP dataset with problems (문제).

```{r}
#| label: messy-data-setup
#| echo: false

library(tidyverse)

set.seed(42)

patients <- tibble(
  person_id = 1:12,
  sex_raw = c("F", "Female", "M", "male", "F ", "Male",
              "f", "MALE", "Female", "M", "Unknown", "female"),
  age = c(45, 62, 71, 38, 55, NA, 29, 84, 150, 66, 41, 58),
  race_raw = c("White", "WHITE", "Black or African American", "Asian",
               "black", "white", "Other", "asian", "White", NA,
               "Black or African American", "Other")
)

patients
```

::: {.callout-important}
What are the problems you see?
:::

---

## And Some Messy Vitals (Wide Format)

```{r}
#| label: messy-vitals-setup
#| echo: false

set.seed(42)

vitals_wide <- tibble(
  person_id = 1:50,
  visit_date = as.Date("2023-01-01") + sample(0:300, 50),
  sbp = sample(80:220, 50, replace = TRUE),
  dbp = sample(40:100, 50, replace = TRUE),
  hr = sample(50:120, 50, replace = TRUE)
)

i <- sample(1:50, 10)
vitals_wide$sbp[i] <- NA_integer_
vitals_wide$sbp[5] <- 999L
vitals_wide$dbp[i] <- NA_integer_
vitals_wide$hr[7] <- 0L

vitals_wide
```

::: {.callout-important}
What are the problems you see?
:::

---

# Part 2: Recoding Data

---

## The Problem: One Concept, Many Spellings

```{r}
#| label: sex-raw-counts
patients |>
  count(sex_raw)
```

12 patients, but **10 distinct values** for a variable that should have 2–3 categories.

---

## Recoding with `case_when()`

`case_when()` is a flexible way to recode — evaluates conditions in order, top to bottom:

```{r}
#| label: recode-case-when
patients <- patients |>
  mutate(
    sex_raw_trimmed = str_trim(sex_raw),
    sex = case_when(
      str_to_lower(sex_raw_trimmed) %in% c("f", "female") ~ "Female",
      str_to_lower(sex_raw_trimmed) %in% c("m", "male")   ~ "Male",
      TRUE ~ NA_character_
    )
  )

patients |>
  count(sex_raw, sex)
```

Notice `"Unknown"` correctly falls through to `NA` — we didn't have to list every bad value, just the good ones.

---

## Recoding with `fct_collapse()`

When you're working with a `factor` and renaming levels, `forcats` functions are better:

```{r}
#| label: recode-fct-collapse
patients <- patients |>
  mutate(
    race_clean = str_to_lower(str_trim(race_raw)),
    race = fct_collapse(
      race_clean,
      "White" = c("white"),
      "Black or African American" = c("black or african american", "black"),
      "Asian" = c("asian"),
      "Other" = c("other")
    )
  )
count(patients, race_raw, race)
```

::: {.callout-note}
`fct_collapse()` is good when grouping many raw labels into a few clean categories; `case_when()` is better when for complex logic or more columns at once.
:::

---

## Recoding Implausible Values (비현실적인 값)

Implausible numbers often need to become missing values (`NA`):

```{r}
#| label: recode-implausible-age
patients <- patients |>
  mutate(age_clean = if_else(age > 0 & age <= 120, age, NA_real_))

patients |>
  select(person_id, age, age_clean)
```

::: {.callout-important}
Never just delete or silently "fix" implausible values — always keep the original column (`age`) alongside the cleaned one (`age_clean`) so your cleaning logic can be checked easily (**reproducible**).
:::

---

# Part 3: Reshaping Data (Pivoting)

---

## Wide vs. Long: Why It Matters

- **Wide format**: one row per patient/visit, one column per variable (`sbp`, `dbp`, `hr`) — good for reading (like a "flowsheet"), may not be good for many `ggplot2`/modeling functions
- **Long format**: one row per patient-variable combination — required for faceted plots, mixed models, and many `tidyverse` summary functions

Some raw EHR data start as **wide**; most analysis and graphing functions need **long**.

---

## `pivot_longer()`: Wide → Long

```{r}
#| label: pivot-longer-example
vitals_long <- vitals_wide |>
  pivot_longer(
    cols = c(sbp, dbp, hr),
    names_to = "vital_sign",
    values_to = "value"
  )

vitals_long |>
  arrange(person_id) |>
  head(10)
```

Each patient-visit now has **3 rows** instead of 1 — one per vital sign.

---

## Show the Plot from Long Data

```{r}
#| label: plot-long-data

ggplot(vitals_long, aes(x = visit_date, y = value, color = vital_sign)) +
  geom_point()
```

::: {.callout-note}
Do these values look ok to you?
:::

---

## `pivot_wider()`: Long → Wide

Sometimes you need to go the other direction: preparing a table for a regression model, where each row must be one patient-visit with separate columns:

```{r}
#| label: pivot-wider-example
vitals_back_to_wide <- vitals_long |>
  pivot_wider(
    names_from = vital_sign,
    values_from = value
  )

identical(arrange(vitals_wide, person_id), arrange(vitals_back_to_wide, person_id))
```

::: {.callout-note}
`pivot_longer()` and `pivot_wider()` are inverses of each other
:::

---

## A Realistic Long-Format: Labs Over Time

```{r}
#| label: labs-long-example
labs_long <- tibble(
  person_id = rep(1:3, each = 4),
  visit_num = rep(1:2, each = 2, times = 3),
  loinc_name = rep(c("Glucose", "Creatinine", "Glucose", "Creatinine"), 3),
  value = c(140, 1.0, 160, 1.1,   95, 0.9, 210, 1.3,   130, 1.2, 145, 1.4)
)

labs_long
```

## Now Wide-Format: Labs Over Time

```{r}
#| label: labs-wide-pivot
labs_wide <- labs_long |>
  pivot_wider(names_from = loinc_name, values_from = value)

labs_wide
```

This is common: labs are long (one row per result), but a model predicting an outcome from glucose *and* creatinine needs them side by side, per visit.

---

# Part 4: Detecting Outliers

---

## Visual Detection: Boxplots

```{r}
#| label: outlier-boxplot
#| code-fold: true
#| fig-width: 7
#| fig-height: 4

ggplot(vitals_wide, aes(x = "", y = sbp)) +
  geom_boxplot(fill = "#2C6E91", alpha = 0.5) +
  geom_jitter(width = 0.05, alpha = 0.6) +
  labs(title = "Systolic Blood Pressure — Outlier Check", x = NULL, y = "SBP (mmHg)") +
  theme_minimal(base_size = 13)
```

The points near 1000 are likely in error (perhaps a misplaced decimal) and not real.

---

## Statistical Detection: Z-Scores

```{r}
#| label: zscore-outlier-detection
#| code-fold: true
vitals_wide |>
  mutate(
    hr_z = (hr - mean(hr, na.rm = TRUE)) / sd(hr, na.rm = TRUE),
    hr_outlier = abs(hr_z) > 3
  ) |>
  select(person_id, hr, hr_z, hr_outlier)
```

Z-scores work well for roughly normal distributions but are sensitive to the extreme values themselves inflating the mean/SD — consider IQR as an alternative.

---

## Domain Knowledge

The most reliable "outlier detector" in clinical data is often a **plausibility range**:

```{r}
#| label: plausibility-flags
vitals_wide |>
  mutate(
    sbp_implausible = sbp < 60 | sbp > 260,
    hr_implausible  = hr < 30 | hr > 220
  ) |>
  select(person_id, sbp, sbp_implausible, hr, hr_implausible)
```

::: {.callout-important}
A statistically "normal" value can appear clinically impossible (e.g., DBP = 0), and a statistical "outlier" can be a real value (e.g., a hypertensive crisis). Use both lenses together, and never auto-delete outliers without documenting the rule you used.
:::

---

# Part 5: Patterns of Missingness (결측)

---

## Why *Patterns* of Missingness (결측) Matter

::: {.incremental}
- Missing data isn't just "how much" — it's "**which values, and why**"
- **MCAR** (Missing Completely At Random) — missingness unrelated to any value, observed or not (rare in real EHR data)
- **MAR** (Missing At Random) — missingness explained by *other observed* variables (e.g., labs missing more often for outpatient than inpatient visits, so encounter type explains missingness)
- **MNAR** (Missing Not At Random) — missingness related to the *unobserved value itself* (e.g., a lab measure of alcohol or drug screen is only present when clinician thinks needed)
:::

::: {.callout-important}
EHR data is very often **MAR or MNAR**, not MCAR — visit type, site, and even disease severity can affect whether something was measured at all.
:::

---

## Make a Big Vital Sign Dataset

```{r}
#| label: make-big-vital-sign
#| echo: false
n <- 100

vitals_wide_big <- tibble(
  person_id = 1:n,
  visit_date = as.Date("2023-01-01") + sample(0:365, n, replace = TRUE),
  sbp = rnorm(n, mean = 128, sd = 15),
  dbp = rnorm(n, mean = 80, sd = 10),
  hr  = rnorm(n, mean = 75, sd = 12)
)

bp_missing_idx <- sample(1:n, size = 15)  # ~15% of visits missing BP entirely

vitals_wide_big <- vitals_wide_big |>
  mutate(
    sbp = if_else(person_id %in% bp_missing_idx, NA_real_, sbp),
    dbp = if_else(person_id %in% bp_missing_idx, NA_real_, dbp)
  )

hr_missing_idx <- sample(setdiff(1:n, bp_missing_idx), size = 10)

vitals_wide_big <- vitals_wide_big |>
  mutate(hr = if_else(person_id %in% hr_missing_idx, NA_real_, hr))

vitals_wide_big
```

---

## Visualizing Missingness with `naniar`

```{r}
#| label: naniar-vis-miss
#| eval: true
#| fig-width: 7
#| fig-height: 4

# install.packages("naniar")
library(naniar)

vitals_wide_big |>
  vis_miss()
```

`vis_miss()` gives a quick heatmap-style overview: which columns have missing data, and roughly how much. Are there any patterns?

---

## Summarizing Missingness by Variable

```{r}
#| label: naniar-miss-var-summary
#| eval: true

vitals_wide_big |>
  miss_var_summary()
```

This gives exact counts and percentages per column.

---

## Visualizing Missingness Patterns Together

```{r}
#| label: naniar-upset
#| eval: true
#| warning: false

library(naniar)

# shows which COMBINATIONS of variables tend to be missing together
vitals_wide_big |>
  gg_miss_upset()
```

Useful when multiple variables are missing — reveals whether missingness occurs **together** vs. **independently**.

---

## VIM Package for Missingness: histMiss

```{r}
#| label: vim-hist

library(VIM)

histMiss(vitals_wide_big[, c("sbp", "dbp", "hr")], only.miss = FALSE, interactive = FALSE)
```

---

## What NOT to Do (하지 마세요)

::: {.incremental}
- **Don't** silently drop all rows with any missing value (`na.omit()`/`drop_na()`) without first checking *how much* data that removes and *whether* it's random
- **Don't** assume a missing lab means "normal" or impute (replace with) a placeholder value without justification
- **Don't** treat 0, 999, -1, etc. as if they were real numeric values — always check for these before summarizing
:::

---

## Recap

::: {.incremental}
- **Recoding** (`case_when()`, `fct_collapse()`, `recode_values()`) turns inconsistent raw categories into clean, analyzable ones — always keep the raw column alongside the cleaned one
- **Reshaping** (`pivot_longer()` / `pivot_wider()`) moves data between the shape it arrives in and the shape your analysis needs — they're inverses of each other
- **Outliers** should be flagged using both statistical rules (e.g., z-score) *and* domain plausibility — neither alone is sufficient
- **Missingness** has *patterns*, not just a percentage — check whether it relates to other variables before deciding how to handle it (`naniar`)
:::


