---
title: "Homework: Recode, Check, Pivot, Fix"
subtitle: "Healthcare Big Data Analytics"
author: "이름"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## The Dataset

We'll use data from:

https://archive.ics.uci.edu/dataset/296/diabetes+130-us+hospitals+for+years+1999-2008

DOI: 10.24432/C5230J

The dataset represents ten years (1999-2008) of clinical care at 130 US hospitals and integrated delivery networks. It includes over 50 features representing patient and hospital outcomes. Information was extracted from the database for encounters that satisfied the following criteria.

(1)	It is an inpatient encounter (a hospital admission).
(2)	It is a diabetic encounter, that is, one during which any kind of diabetes was entered into the system as a diagnosis.
(3)	The length of stay was at least 1 day and at most 14 days.
(4)	Laboratory tests were performed during the encounter.
(5)	Medications were administered during the encounter.

The data contains such attributes as patient number, race, gender, age, admission type, time in hospital, medical specialty of admitting physician, number of lab tests performed, HbA1c test result, diagnosis, number of medications, diabetic medications, number of outpatient, inpatient, and emergency visits in the year before the hospitalization, etc.

```{r}
#| label: hw-data-setup
#| eval: false

# After unzipping the downloaded file, read in "diabetic_data.csv" for use 
```

---

## Your Tasks

1. **Recode** — `race` has "?" for missing, and `gender` has "Unknown/Invalid". Recode them into `NA` as separate columns. 

```{r}
#| label: hw-data-recode

# Recode data here...
```


2. **Check missingness** — Using `naniar`, `VIM`, or your own code, check how much data is missing for selected variables (no need to do all, such as medications). Is missingness higher for one variable than the others?

```{r}
#| label: hw-data-missing

# Check for missingness in the data here...
```

3. **Pivot** — Diagnoses are wide (`diag_1`, `diag_2`, `diag_3`). Create a long dataset (using `pivot_longer`) that includes only the `patient_nbr` and `diagnosis` (combining the 3 diagnosis columns into one). Recode missing "?" as `NA`. What is the average number of *unique* diagnoses per patient (ignoring `NA`, across all encounters, and do not use `number_diagnoses` which is the number diagnoses entered into the system)? 

```{r}
#| label: hw-data-pivot

# Pivot data into long format and recoding missing here...
```

4. **Check for outliers** — Check distributions and for outliers using `boxplot` or alternative code, specifically on `num_lab_procedures`, `num_medications`, `number_emergency`, and `number_outpatient`. For these variables, describe if you believe any outliers are in error or not, and why.

```{r}
#| label: hw-data-outliers

# Check distributions and outliers in the data here...
```

---

## Submission

- Render to `.html` (`quarto render`) showing all code, output, and brief written explanations as comments
- Code must run with **Restart R and Run All** before submitting

