---
title: "Homework: Grouping Diagnoses in a Mock OMOP CDM"
subtitle: "Big Medical & Health Data with R"
author: "이름"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

## Assignment Overview

**Goal:** Practice grouping clinical facts by diagnosis using a fake OMOP CDM, then verify one diagnosis code's hierarchy in **Athena**.

You will:

1. Build a mock OMOP CDM in R (`omock` + `CDMConnector`)
2. Group `condition_occurrence` records by diagnosis (`concept_name`)
3. Look up one diagnosis in Athena and find its descendant concepts
4. Re-run your grouping using the full descendant set (a mini phenotype!)

## Step 0 — Setup

Note data is less than 1 GB download.

```{r}
#| label: setup
# install.packages(c("omock", "CDMConnector", "dplyr"))
library(omock)
library(CDMConnector)
library(dplyr)

# Build a mock CDM from the synthea-heart-10k fake dataset
cdm <- mockCdmFromDataset(datasetName = "synthea-heart-10k", source = "duckdb")

cdm$condition_occurrence
cdm$concept
```

`cdm` behaves like a named list of tables (`person`, `condition_occurrence`, `concept`, ...).

## Task 1 — Group by Diagnosis

Join `condition_occurrence` to `concept` on `condition_concept_id = concept_id`, then count records **and** distinct patients per diagnosis. Order results by most frequent conditions.

```{r}
#| label: task1-starter
diagnosis_counts <- cdm$condition_occurrence |>
  left_join(
    cdm$concept,
    by = c("condition_concept_id" = "concept_id")
  ) |>
  group_by(concept_name) |> 
  summarise( # can also use reframe
    n_records  = n(),
    n_patients = n_distinct(person_id)
  ) |>
  collect()

diagnosis_counts |>
  arrange(desc(n_records))
```

> **Q1:** Which diagnosis has the most records? Does it also have the most distinct patients? Why might those two rankings differ in other data sources?

## Task 2 — Athena + Hierarchy

1. Pick **one** diagnosis from your Task 1 table
2. Look it up on <https://athena.ohdsi.org> — search the `concept_name`
3. Record its `concept_id`, `vocabulary_id`, and `standard_concept` flag
4. Open its **"View Details" → Relationships/Hierarchy** tab and list 2–3 **descendant** concepts

```{r}
#| label: task2-starter
# TODO: replace with the concept_id you found on Athena, an "ancestor" that has descendants (use cdm$concept_ancestor for possible choices)
my_concept_id <- 441840

descendants <- cdm$concept_ancestor |>
  filter(ancestor_concept_id == my_concept_id) |>
  pull(descendant_concept_id)

# Re-run Task 1's grouping, but now include the diagnosis AND all its descendants
cdm$condition_occurrence |>
  filter(condition_concept_id %in% c(my_concept_id, descendants)) |>
  summarise(n_records = n(), n_patients = n_distinct(person_id)) |>
  collect()
```

> **Q2:** Are the n_records and n_patients different now? Why? And what does that imply for a real phenotype definition that only used one exact code?

