---
title: "Healthcare Big Data Analytics"
subtitle: "Introduction"
author: "Ben Gerber, MD, MPH"
date: last-modified
format:
  revealjs:
    smaller: true
    theme: dracula
    slide-number: true
    footer: "Healthcare Big Data Analytics — Introduction"
    toc: false
    incremental: false
    code-fold: false
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

library(iglu)
library(medicaldata)
library(EHR)
library(pkdata)
library(tidyverse)
```


## Welcome (환영합니다)

::: {.incremental}
- Instructors: Profs. Dong-soo Shin and Ben Gerber
- Contact: shindong@hallym.ac.kr / ben.gerber@umassmed.edu
:::

---

## Introductions

Please share:

1. Program / department
2. Prior work with **R** or coding (none / some / a lot)
3. Prior work with clinical, electronic health record (EHR), or health data
4. One thing you hope to learn with our class
5. Something **fun** you like

---

## Why Healthcare Big Data?

- Lot of EHR (전자건강기록) use and growth of wearable/sensor data
- Real-world data (RWD) used for clinical, policy, and regulatory decisions
- Need for researchers who can do:
  - Clinical data → reproducible analysis → important knowledge

---

## Course Description

> This course provides the basic knowledge and skills to do research with healthcare data — EHRs, wearable device data, and other real-world health data.

By the end, you will be able to:

- Do medical data-driven analysis
- Manage, clean, and analyze health data in R
- Understand some clinical coding systems and data standards
- Apply ethical and reproducible research practices

---

## Course Lessons — Roadmap

::: {.fragment .smaller}
1. Introduction, research questions, data sources, and tools
2. Medical data structures and relationships
3. Data cleaning, wrangling, reshaping, missingness
4. Coding systems and phenotypes
5. Cohort construction
6. Exploratory data analysis
7. Longitudinal data with wearables
8. Statistical modeling
9. Machine learning, prediction
10. Ethics, privacy, reproducibility
:::

---

## Tools We'll Use

::: {.columns}
::: {.column width="50%"}
**R Language**

- [**R Project**](https://www.r-project.org/) 
- [**R Studio**](https://posit.co/downloads) as IDE
:::

::: {.column width="50%"}
**Reproducibility**

- [**Quarto**](https://quarto.org/) for reproducible reports
- [**git**](https://git-scm.com/) for version control
- [**GitHub**](https://github.com/) account for collaboration
:::
:::

---

## Workflows for Reproducibility

https://hbiostat.org/rflow/

```{mermaid}
%%| fig-width: 8
%%| eval: true
flowchart TD
  R[R Workflow] --> Rformat[Report formatting]
  Rformat --> Quarto["Quarto setup, metadata,<br>table & graph formatting"]
  R --> DI[Data import] --> Annot["Annotate data<br>(labels, units, data dictionary)"]
  R --> Do[Data overview] --> F["Observation filtration,<br>missing data patterns"]
  R --> P[Data processing] --> DP["Recode / transform / reshape /<br>merge / aggregate"]
  R --> Des["Descriptive statistics<br>(univariate, stratified)"]
  R --> An["Analysis<br>(stay close to the data)"]
```

---

## Example De-Identified EHR/ICU (중환자실) Datasets {.scrollable}

| Dataset | Source | Notes |
|---|---|---|
| **MIMIC-IV** | Beth Israel Deaconess (Boston, USA) | Hospital admissions 2008–2019 |
| **eICU-CRD** | Multi-center (USA) | 200,000+ ICU admissions |
| **AmsterdamUMCdb** | Netherlands | 1st open European ICU database; now in OMOP |
| **HiRID** | Switzerland | High time-resolution; ~34k patients |
| **BlendedICU** | Harmonized (Multiple) | 309k+ stays, in OMOP |
| **CRITICAL** | Multi-site US | 1.95B records, 371K patients, OMOP |

---

## What EHR Data Looks Like After Wrangling

```{r}
#| label: example-ehr-data-preview

# define directories and read sample data
checkDir <- file.path(tempdir(), 'check1')
rawDataDir <- system.file("examples", "str_ex1", package="EHR")
if (!dir.exists(checkDir)) dir.create(checkDir)
demo <- read.csv(file.path(rawDataDir,"Demographics_DATA_simple.csv"))
head(demo, c(6, 7))
```

---

## What Wearable Data Looks Like (Glucose Monitoring)

```{r}
#| label: wearable-preview
#| fig-width: 9
#| fig-height: 4.5

head(example_data_1_subject)
plot_glu(example_data_1_subject, plottype = 'tsplot', tz = "EST")
```

---

## Clinical Coding Systems

| System | Used For |
|---|---|
| **ICD** and **KCD** | Diagnoses |
| **CPT** and **EDI** | Procedures / billing |
| **LOINC** and **K-LOINC** | Lab & clinical observations |
| **RxNorm** | Medication vocabularies | 
| **SNOMED CT** | Clinical terminology / findings |
| **OMOP CDM** | Common data model |

---

## Prerequisites (선수 과목)

- Basic R (variables, data frames, functions) *or* willingness to catch up quickly :)
- Introductory statistics (regression, hypothesis testing)
- No prior clinical/health data experience required

::: {.callout-note}
No worries about your R experience - we can help!
:::

---

## Grading Breakdown

```{r}
#| label: grading-pie
#| echo: false
#| fig-width: 8
#| fig-height: 5

grading <- data.frame(
  component = c("Homework", "Final Project", "Attendance"),
  weight = c(40, 40, 20)
)

ggplot(grading, aes(x = "", y = weight, fill = component)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(weight, "%")),
            position = position_stack(vjust = 0.5), color = "white", fontface = "bold", size = 12) +
  scale_fill_manual(values = c("#2C6E91", "#E07A5F", "#81B29A")) +
  theme_void() +
  theme(legend.title = element_blank(), legend.text = element_text(size = 24), legend.position = "right")
```

---

## Grading Details

**Homework and Process Monitoring — 40%**

- Weekly homework with R
- Final project check-ins

**Final Project Report and Presentation — 40%**

- Full workflow: research question → cohort → analysis → reproducible report
- Rubric: research question clarity, analytic rigor, reproducibility, presentation quality

**Attendance — 20%**

- Active participation counts, not just physical presence

---

## Final Project

The capstone of this course: apply the **complete workflow** to a healthcare research question.

::: {.incremental}
1. Formulate a research question
2. Construct a cohort from EHR / wearable / public data
3. Clean, explore, and analyze the data
4. Apply model
5. Produce a reproducible report (Quarto + GitHub) including background, aims, methods, results, discussion, limitations
6. Present findings to the class
:::

**Milestones:** Proposal (Week ~4) → Analysis Checkpoint (Week ~8) → Final Report & Presentation (TBD)

---

## From Curiosity to a Research Question (연구 질문)

Big medical data doesn't start with data — it starts with a **question worth answering**.

::: {.incremental}
- A dataset with 50M patients can't help a vague or unanswerable question
- A common statement: *"I have access to X data, what can I do with it?"* — start from question
:::

---

## Where Good Questions Come From

::: {.columns}
::: {.column width="50%"}
**Clinical / practical sources**

- A pattern you've noticed in practice or reading
- A guideline with weak evidence behind it
- Conflicting findings across prior studies
- A subgroup consistently excluded from trials (elderly or pregnant patients)
:::

::: {.column width="50%"}
**Data-driven sources**

- A gap: no one has studied *this* outcome in *this* population
- A new data source becoming available (e.g., wearables)
- Replication: does a finding from Population A hold in Population B?
:::
:::

---

## The FINER Criteria

A checklist for making a research question:

| Criterion | Ask yourself |
|---|---|
| **F**easible | Enough patients, follow-up time, and data quality to answer it? |
| **I**nteresting | Would you want to know the answer? |
| **N**ovel | Does it confirm, refute, or extend what's already known? |
| **E**thical | Can it be done without too much risk or privacy violation? |
| **R**elevant | Does it matter for clinical practice, policy, or future research? |

---

## Structuring the Question: PECO

Useful for comparative or causal questions:

| Element | Meaning | Example |
|---|---|---|
| **P** | Population | Adults newly diagnosed with type 2 diabetes |
| **E** | Exposure | Initiation of SGLT2 inhibitor |
| **C** | Comparison | Initiation of metformin monotherapy |
| **O** | Outcome | Hospitalization for heart failure within 2 years |

**Resulting question:** *"Among adults newly diagnosed with type 2 diabetes, does starting an SGLT2 inhibitor (vs. metformin) lower the risk of heart-failure hospitalization?"*

---

## Not All Questions Are Causal

Big medical data supports several distinct question *types*:

::: {.incremental}
- **Descriptive** — "What is the prevalence of X in population Y?"
- **Predictive** — "Can we predict which patients will develop X within N days?"
- **Comparative effectiveness** — "Does treatment A lead to better outcomes than treatment B?"
- **Health service use** — "How does access to Y vary by Z (e.g., demographics)?"
:::

::: {.callout-important}
Confusing *prediction* with *causal inference* is a common error — a model that predicts risk well does not tell you what to change to reduce it.
:::

---

## Strengths (장점) of Medical Data

::: {.incremental}
- **Scale** — millions of patients enable studying rare outcomes or small subgroups
- **Real-world generalizability** — captures patients excluded from RCTs (elderly, non-adherent)
- **Longitudinal** — years of follow-up, sometimes across a full lifespan
- **Speed & cost** — no new data collection required; faster than prospective studies
- **Naturally occurring variation** — comparative effectiveness research that would be unethical or impractical to randomize
- **Granularity** — wearables/sensors provide continuous physiological data unavailable in clinic visits
:::

---

## Limitations (제한점) of Medical Data

::: {.incremental}
- **Confounding by indication** — sicker patients get certain treatments (bias)
- **Missing data** — not missing at random; care received *outside* the system is invisible
- **Measurement error** — codes reflect billing/documentation practices, not always clinical truth
- **Selection bias** — who's *in* the dataset is rarely a random sample of the population of interest
- **Immortal time bias** — common pitfall in EHR cohort studies (e.g., statin post-MI)
- **Generalizability** — single-institution or single-country data may not transport elsewhere
- **Data heterogeneity** — coding systems and practices shift across sites, time, and countries
:::

---

## Ethics (윤리), Privacy & Data Governance

- De-identification and date shifting
- Human subject considerations for secondary use of health data
- Data use agreements (DUA) when working with real EHR/wearable data
- Why we use **synthetic data** in this course — realistic structure, no privacy risk

::: {.callout-important}
These considerations aren't an afterthought — they're a required part of homework and final project.
:::

---

## Academic Integrity & AI-Use Policy

- Collaboration is encouraged for learning; but submitted work must be your own
- AI (Copilot, ChatGPT, Claude, Posit Assistant, etc.):
  - Allowed for debugging/syntax help, **must be disclosed**, not allowed for full assignment generation.
  - AI-assisted portions of code must be disclosed briefly at the top of your script. 
  - You are responsible for understanding, and able to explain, any code you submit.
- All final project code must run reproducibly

---

## Questions?

질문 있으세요?

---

## Installing R

https://www.r-project.org/

---

## Installing R Studio

https://posit.co/downloads

---

## Installing Git

https://git-scm.com/install/

---

## Getting Account on GitHub

https://github.com/signup

---

## Installing Packages

```{r}
install.packages("tidyverse", repos = "http://cran.us.r-project.org")
install.packages("medicaldata", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(medicaldata)

glimpse(scurvy)
```

## Getting Familiar with R Studio

If time available in class, we can review the basics.

