---
title: "OMOP CDM and Cohort Creation"
subtitle: "Going Deep into OMOP"
author: "Ben Gerber, MD, MPH"
date: last-modified
format:
  revealjs:
    theme: dracula
    slide-number: true
    footer: "Big Medical Data with R"
    chalkboard: true
    smaller: true
    code-line-numbers: true
    incremental: false
    width: 1280
    height: 720
execute:
  echo: true
  warning: false
  message: false
---

## Today's Agenda

1. Creating a CDM reference
2. Exploring OMOP CDM
3. Identifying patient characteristics
4. Adding cohorts to the CDM
5. Working with cohorts

# 1. Creating a CDM reference

## OMOP CDM

The OMOP CDM standardizes the structure of healthcare data. Data is stored across a system of tables with established relationships between them (relational database / 관계형 DB).

Here is version 5.4 of OMOP CDM (Version 5.5 was released in August 2026):

https://ohdsi.github.io/CommonDataModel/cdm54erd.html

## Reference to the OMOP CDM

- We will create a data model in R to represent the OMOP CDM.
- This can provide a basis for analytic pipelines using the data.
- Our `cdm` object contains references to all the OMOP CDM tables (note version of this test CDM).

Note that here we have first created a local version of the `cdm` with all the tables of interest with `omock` (`cdm_local`), then copied it to a `DuckDB` database, and finally created a reference to it with `CDMConnector`, so that we can work with the final `cdm` object as we normally would for one created with our own healthcare data. This is a database-backed reference for fast/optimized SQL queries (similar to real OMOP instance).

```{r}
#| label: create-cdm-reference
#| code-fold: true

library(duckdb)
library(tidyverse)
library(omock)
library(CDMConnector)
library(omopgenerics)
library(CohortConstructor)
library(CohortCharacteristics)

cdm_local <- mockCdmReference() |>
  mockPerson(nPerson = 100) |>
  mockObservationPeriod() |>
  mockConditionOccurrence() |>
  mockDrugExposure() |>
  mockObservation() |>
  mockMeasurement() |>
  mockVisitOccurrence() |>
  mockProcedureOccurrence()

con <- dbConnect(drv = duckdb())
src <- dbSource(con = con, writeSchema = "main")

cdm <- insertCdmTo(cdm = cdm_local, to = src)

cdm

cdm$cdm_source |>
  pull(cdm_version)
```

## Accessing a Table

- You can access specific tables using the “$” or “[[ … ]]” operators (extraction). Did you know single and double brackets are functions (함수)?

The `person` table has demographic and identity information (1:1 relationship with patient).

The `observation_period` defines a time window when clinical data is complete and available (when a patient is in observation).

```{r}
#| label: accessing-table

cdm$person

cdm[["observation_period"]]
```

# 2. Working with Mock Data

## Use Mock COVID-19 Data

We'll try using the COVID-19 fake dataset. This data can be relatively large in size.

```{r}
#| label: get-covid-data

cdm <- mockCdmFromDataset(datasetName = "synthea-covid19-10k", source = "duckdb")

cdm
```

## Counting People

How many people are in this dataset? And how many by gender?

```{r}
#| label: counting-people
#| code-fold: true

cdm$person |>
  count()

cdm$person |>
  group_by(gender_concept_id) |>
  count() |>
  left_join(cdm$concept, by = c("gender_concept_id" = "concept_id")) |>
  select("gender_concept_id", "concept_name", "n") |>
  collect()
```


## Summarizing Observation Periods

- The observation period table contains records showing spans of time over which clinical events are observed for the people in the person table. 
- Someone can potentially have multiple observation periods. 

Let's count number of people grouped by the year during which their first observation period started.

```{r}
#| label: summarize-observation-periods
#| code-fold: true

first_observation_period <- cdm$observation_period |>
  group_by(person_id) |>
  arrange(observation_period_start_date) |>
  filter(row_number() == 1) |> # can also use "slice"
  compute() # keep in db as opposed to collect()

# add to the person table
first_records_per_year <- cdm$person |>
  left_join(first_observation_period, by = "person_id") |>
  mutate(observation_period_start_year = get_year(observation_period_start_date)) |>
  group_by(observation_period_start_year) |>
  count() |>
  collect()

tail(first_records_per_year)
```

## First Year Observations Plotted

```{r}
#| label: first-year-observations-plot
#| code-fold: true

ggplot(first_records_per_year) +
  geom_col(mapping = aes(x = observation_period_start_year, y = n)) +
  theme_bw()

```

## Counts of COVID Symptoms

For a group of codes, get counts of those COVID symptoms. (For example, 437663 is fever - can you verify?)

```{r}
#| label: counts-covid-symptoms
#| code-fold: true

codes <- c(
  437663, 437390, 31967, 4289517, 4223659,
  312437, 434490, 254761, 77074
)

symptoms_records <- cdm$condition_occurrence |>
  filter(condition_concept_id %in% codes) |>
  group_by(condition_concept_id) |>
  count() |>
  left_join(cdm$concept |>
              select(condition_concept_id = concept_id, condition_concept_name = concept_name),
            by = "condition_concept_id"
  ) |>
  collect()

ggplot(symptoms_records) +
  geom_col(mapping = aes(x = condition_concept_name, y = n)) +
  theme_bw() +
  xlab("Symptom")
```

# 3. Adding OMOP CDM Cohorts

## What is a cohort (코호트)?

- When conducting research using the OMOP CDM, we may want to identify groups of people who share specific characteristics. 
- The inclusion criteria for these groups can range from simple to complex.

**Simple:** diagnosed with asthma

**Complex:** diagnosed with asthma AND:

- adult age 18+
- at least 1 year of observation before diagnosis
- no prior history of chronic obstructive pulmonary disease (COPD)
- no history of using short-acting beta-antagonists medication

## OMOP Cohort

The groups of people we identify are called cohorts. In the OMOP CDM, cohorts use a specific structure: a cohort table with four required fields:

1. `cohort_definition_id` a unique ID for each cohort (multiple cohorts can be defined in the same cohort table).
2. `subject_id` a foreign key linking subject to the person table.
3. `cohort_start_date` date indicating the beginning of the cohort record.
4. `cohort_end_date` date indicating the end of the cohort record.

Individuals must be defined in the person table and under observation (i.e. have an ongoing record in the observation period table). 

Individuals can enter a cohort multiple times, but the time periods cannot overlap.

## Atlas

ATLAS is an OHDSI tool to conduct analyses with OMOP CDM. 

https://atlas-demo.ohdsi.org

You can create cohorts by defining groups of people based on an exposure to a drug or diagnosis. 

**Concept Sets:** allow for groups of `concept_id`s to define clinical concepts for a research question and can be reused and shared.

**Cohort Definitions:** create a set of persons who satisfy criteria over time, used in analyses, and can be defined based on entry event (first occurrence or inclusion criteria)

**Characterizations:** analytic capability to look at 1 or more cohorts and summarize characteristics

## COVID Cohort

Using the `CohortConstructor` package we can make a cohort of people with COVID-19.

Atlas will help us find a concept set to use:

https://atlas-demo.ohdsi.org/#/concept/37311061

```{r}
#| label: make-covid-cohort

cdm$covid <- conceptCohort(
  cdm = cdm,
  conceptSet = list("covid" = 37311061),
  name = "covid"
)

cdm$covid
```

## Apply Inclusion Criteria

What if we want to limit by dates? Or demographics?

```{r}
#| label: apply-inclusion-criteria
#| code-fold: true

cdm$covid <- cdm$covid |>
  requireInDateRange(dateRange = c(as.Date("2021-01-01"), NA)) 

cdm$covid <- cdm$covid |>
  requireDemographics(ageRange = c(18, 64), sex = "Male") 

cdm$covid |>
  count()
```

## Summary of Inclusion/Exclusion Criteria

```{r}
#| label: summary-criteria
#| code-fold: true

attrition_summary <- summariseCohortAttrition(cohort = cdm$covid)
tableCohortAttrition(attrition_summary)
```

## Close Up!

```{r}
#| label: close-up

cdmDisconnect(cdm)
```





